---
title: "Gapminder Exploration"
author: "Kenny Chiu"
date: "2019-09-17"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gapminder)
library(dplyr)
library(ggplot2)
df = gapminder
```

## Introduction

We explore the `gapminder` dataset that is available from the `gapminder` R package. The dataset has `r nrow(df)` observations and has `r ncol(df)` variables including

* country
* continent
* year
* life expectancy at birth
* total population
* per-capita GDP (in units of international dollars)


## Introduction

The dataset includes observations from `r length(levels(df$continent))` continents. The number of countries included in the dataset is shown below.

```{r, echo=FALSE}
df[c("country", "continent")] %>%
  distinct() %>%
  group_by(continent) %>%
  summarise(num_country = length(continent)) %>%
  ggplot(aes(x = continent, y = num_country)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = num_country), vjust = -0.3) +
    labs(title = "Number of Countries by Continent", x = "Continent", y = "Number of Countries") +
    theme(plot.title = element_text(hjust = 0.5))
```


## Population by Continent

We visualize the mean country population per year for each continent/region. The mean population appears to be increasing over time in each continent.

```{r, echo=FALSE, fig.width=8, fig.height=4}
df[c("continent", "year", "pop")] %>%
  group_by(continent, year) %>%
  summarise(mean_total_pop = mean(pop)) %>%
  ggplot(aes(x = year, y = mean_total_pop, group = continent)) +
    geom_line(aes(color = continent)) +
    labs(title = "Mean Country Population per Year by Continent", x = "Year", y = "Mean Country Population", color = "Continent") +
    theme(plot.title = element_text(hjust = 0.5))
```


## Population by Country

We visualize the total population over time for each country in the Oceania region. The population appears to grow at a faster rate in Australia compared to New Zealand.

```{r, echo=FALSE, fig.width=8, fig.height=4}
df[c("country", "continent", "year", "pop")] %>%
  filter(continent == "Oceania") %>%
  ggplot(aes(x = year, y = pop)) +
    geom_line(aes(color = country)) +
    labs(title = "Oceania Population per Year by Country", x = "Year", y = "Population", color = "Country") +
    theme(plot.title = element_text(hjust = 0.5))
```