---
title: "Gapminder Exploration"
author: "Kenny Chiu"
date: "2019-09-17"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gapminder)
library(dplyr)
library(ggplot2)
df = gapminder
```

## Introduction

We explore the `gapminder` dataset that is available from the `gapminder` R package. The dataset has `r nrow(df)` observations and has `r ncol(df)` variables including

* country
* continent
* year
* life expectancy at birth
* total population
* per-capita GDP (in units of international dollars)

The dataset includes observations from `r length(levels(df$continent))` continents. The number of countries included in the dataset is shown in the following plot.

```{r}
df[c("country", "continent")] %>%
  distinct() %>%
  group_by(continent) %>%
  summarise(num_country = length(continent)) %>%
  ggplot(aes(x = continent, y = num_country)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = num_country), vjust = -0.3) +
    labs(title = "Number of Countries by Continent", x = "Continent", y = "Number of Countries") +
    theme(plot.title = element_text(hjust = 0.5))
```